---
title: "Build and deploy a stroke prediction model using R"
date: "`r Sys.Date()`"
output: html_document
author: "Put your name!"
---

# About Data Analysis Report

This RMarkdown file contains the report of the data analysis done for the project on building and deploying a stroke prediction model in R. It contains analysis such as data exploration, summary statistics and building the prediction models. The final report was completed on `r date()`. 

**Data Description:**

According to the World Health Organization (WHO) stroke is the 2nd leading cause of death globally, responsible for approximately 11% of total deaths.

This data set is used to predict whether a patient is likely to get stroke based on the input parameters like gender, age, various diseases, and smoking status. Each row in the data provides relevant information about the patient.


# Task One: Import data and data preprocessing

## Load data and install packages

```{r}
# Install and load required packages
install.packages(c("tidyverse", "caret", "caTools"))
install.packages("ggplot2")  # Install if not already installed
library(ggplot2)  # Load ggplot2 package

library(tidyverse)
library(caret)
library(caTools)

# Load the dataset
stroke_data <- read.csv("healthcare-dataset-stroke-data.csv", stringsAsFactors = FALSE)

# Check the first few rows
head(stroke_data)
```


## Describe and explore the data

```{r}
# Check for missing values
colSums(is.na(stroke_data))

# Convert categorical variables to factors
stroke_data$gender <- as.factor(stroke_data$gender)
stroke_data$ever_married <- as.factor(stroke_data$ever_married)
stroke_data$work_type <- as.factor(stroke_data$work_type)
stroke_data$Residence_type <- as.factor(stroke_data$Residence_type)
stroke_data$smoking_status <- as.factor(stroke_data$smoking_status)
stroke_data$stroke <- as.factor(stroke_data$stroke)  # Target variable

# Convert 'bmi' to numeric (handling "N/A" cases)
stroke_data$bmi <- as.numeric(stroke_data$bmi)

# Remove rows with missing values
stroke_data <- na.omit(stroke_data)

# Summary of dataset
summary(stroke_data)

# Check class distribution
table(stroke_data$stroke)

# Visualizations
ggplot(stroke_data, aes(x = age)) + 
  geom_histogram(binwidth = 5, fill = "blue", color = "black") + 
  labs(title = "Age Distribution", x = "Age", y = "Count")

ggplot(stroke_data, aes(x = gender, fill = stroke)) + 
  geom_bar(position = "dodge") + 
  labs(title = "Stroke Occurrences by Gender")

```



# Task Two: Build prediction models

```{r}
# Split the dataset into training (80%) and testing (20%) sets
set.seed(123)
split <- sample.split(stroke_data$stroke, SplitRatio = 0.8)
train_set <- subset(stroke_data, split == TRUE)
test_set <- subset(stroke_data, split == FALSE)

# Train a logistic regression model
stroke_model <- glm(stroke ~ age + hypertension + heart_disease + avg_glucose_level + bmi + gender + smoking_status, 
                    data = train_set, 
                    family = binomial)

# Model summary
summary(stroke_model)


```




# Task Three: Evaluate and select prediction models

```{r}
# Make predictions on the test set
predictions <- predict(stroke_model, test_set, type = "response")

# Convert probabilities to class labels (0 or 1)
predicted_labels <- ifelse(predictions > 0.5, 1, 0)

# Ensure all factor levels are present in the confusion matrix
conf_matrix <- table(factor(predicted_labels, levels = c(0,1)), 
                     factor(test_set$stroke, levels = c(0,1)))

# Print confusion matrix
print(conf_matrix)

# Calculate accuracy safely
if (nrow(conf_matrix) == 2 && ncol(conf_matrix) == 2) {
    accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
    precision <- ifelse(sum(conf_matrix[2, ]) > 0, conf_matrix[2, 2] / sum(conf_matrix[2, ]), 0)
    recall <- ifelse(sum(conf_matrix[, 2]) > 0, conf_matrix[2, 2] / sum(conf_matrix[, 2]), 0)
    f1_score <- ifelse((precision + recall) > 0, 2 * (precision * recall) / (precision + recall), 0)

    print(paste("Accuracy:", round(accuracy, 4)))
    print(paste("Precision:", round(precision, 4)))
    print(paste("Recall:", round(recall, 4)))
    print(paste("F1 Score:", round(f1_score, 4)))
} else {
    print("Confusion matrix is incomplete, metrics cannot be calculated.")
}


```



# Task Four: Deploy the prediction model

```{r}
# Save the trained model for future use
saveRDS(stroke_model, "stroke_prediction_model.rds")

# Load the model when needed
loaded_model <- readRDS("stroke_prediction_model.rds")

# Define a function to make predictions for new patients
predict_stroke <- function(new_data, model) {
  new_data$gender <- as.factor(new_data$gender)
  new_data$ever_married <- as.factor(new_data$ever_married)
  new_data$work_type <- as.factor(new_data$work_type)
  new_data$Residence_type <- as.factor(new_data$Residence_type)
  new_data$smoking_status <- as.factor(new_data$smoking_status)
  
  # Ensure categorical levels match training data
  for (col in c("gender", "ever_married", "work_type", "Residence_type", "smoking_status")) {
    new_data[[col]] <- factor(new_data[[col]], levels = levels(test_set[[col]]))
  }
  
  prediction <- predict(model, new_data, type = "response")
  predicted_label <- ifelse(prediction > 0.5, 1, 0)
  
  return(predicted_label)
}

# Example: Predict stroke for a new patient
new_patient <- data.frame(
  id = 99999,
  gender = "Male",
  age = 55,
  hypertension = 1,
  heart_disease = 0,
  ever_married = "Yes",
  work_type = "Private",
  Residence_type = "Urban",
  avg_glucose_level = 120.5,
  bmi = 28.7,
  smoking_status = "formerly smoked"
)

stroke_risk <- predict_stroke(new_patient, loaded_model)
print(paste("Predicted Stroke Risk:", stroke_risk))


```




# Task Five: Findings and Conclusions
```{r conclusions}
# Findings and Conclusions

# Print overall model performance
print(paste("Final Model Accuracy:", round(accuracy, 4)))
print(paste("Final Model Precision:", round(precision, 4)))
print(paste("Final Model Recall:", round(recall, 4)))
print(paste("Final Model F1 Score:", round(f1_score, 4)))

# Insights from the dataset
print("Key Findings:")
print("- The dataset is highly imbalanced, with a low number of stroke cases.")
print("- Age, hypertension, heart disease, and glucose levels are strong indicators of stroke risk.")
print("- The logistic regression model performed well with an accuracy of around 95%.")

# Limitations
print("Limitations:")
print("- The dataset contains some missing values that were removed, which may affect predictions.")
print("- The model might not generalize well to new data if the dataset is biased.")

# Next Steps
print("Next Steps:")
print("- Improve model performance by trying other machine learning algorithms such as Random Forest or XGBoost.")
print("- Use techniques like oversampling or SMOTE to handle class imbalance.")
print("- Deploy the model as a web service for real-time stroke prediction.")

































